\documentclass[11pt]{article}
\newcommand{\courseshort}{Data Structures (Make-up) · 010153103}
\newcommand{\lecturetitle}{L3: Linked Lists}
\usepackage{lecture_style}

\begin{document}
\lectureheader{Lecture 3: Singly \& Doubly Linked Lists}{Pre-class brief}{Goodrich Ch.~3, 7.3--7.4}

\begin{objbox}
\begin{itemize}[leftmargin=*]
  \item Build a node class and link nodes via references.
  \item Implement \texttt{addFirst}, \texttt{addLast}, \texttt{removeFirst}, \texttt{removeLast} on singly and doubly linked lists.
  \item Use sentinel nodes (header/trailer) and explain why they simplify code.
  \item Choose between array-based and linked structures based on operation costs.
\end{itemize}
\end{objbox}

\section*{1. Singly Linked List (SLL)}

\begin{center}
\begin{tikzpicture}[
  node/.style={rectangle split, rectangle split parts=2, rectangle split horizontal,
               draw, thick, minimum height=0.7cm, rectangle split part fill={white, gray!10}},
  ptr/.style={->, thick, blue, >=Stealth}
]
  \node[node] (a) {\textbf{a} \nodepart{two} \phantom{xx}};
  \node[node, right=1.4cm of a] (b) {\textbf{b} \nodepart{two} \phantom{xx}};
  \node[node, right=1.4cm of b] (c) {\textbf{c} \nodepart{two} \phantom{xx}};
  \node[right=1.0cm of c] (null) {$\phi$};
  \node[above=0.7cm of a, draw, circle, fill=red!10] (h) {head};
  \draw[ptr] (h) -- (a.north);
  \draw[ptr] (a.two) -- (b.west);
  \draw[ptr] (b.two) -- (c.west);
  \draw[ptr] (c.two) -- (null.west);
\end{tikzpicture}
\end{center}

\begin{lstlisting}
class Node<E> {
    E element;
    Node<E> next;
    Node(E e, Node<E> n) { element = e; next = n; }
}
\end{lstlisting}

\textbf{Costs.} \texttt{addFirst}, \texttt{removeFirst} are $O(1)$. \texttt{addLast} is $O(1)$ \emph{only} if a tail pointer is kept; otherwise $O(n)$. \textbf{\texttt{removeLast} is always $O(n)$} on an SLL because there is no back pointer.

\section*{2. Doubly Linked List (DLL)}

Each node has both \texttt{prev} and \texttt{next}. With \emph{header} and \emph{trailer} sentinel nodes, every real node has non-null neighbors --- no special cases for the ends.

\begin{lstlisting}
class DNode<E> {
    E element;
    DNode<E> prev, next;
}
public void addBetween(E e, DNode<E> a, DNode<E> b) {
    DNode<E> x = new DNode<>(); x.element = e;
    x.prev = a; x.next = b;
    a.next = x; b.prev = x;
}
\end{lstlisting}

\textbf{Costs (DLL).} \texttt{addFirst}, \texttt{addLast}, \texttt{removeFirst}, \texttt{removeLast} are all $O(1)$.

\section*{3. Array vs.\ Linked --- when to use which?}

\begin{center}
\begin{tabular}{lll}
\textbf{Operation} & \textbf{ArrayList} & \textbf{DLL} \\
\hline
random access \texttt{get(i)}    & $O(1)$ & $O(n)$ \\
add/remove at end (tail)         & $O(1)^*$ & $O(1)$ \\
add/remove at front (head)       & $O(n)$ & $O(1)$ \\
add/remove at known node         & $O(n)$ & $O(1)$ \\
memory per element               & low & higher (2 pointers) \\
\end{tabular}\\[2pt]
\small $^*$amortized
\end{center}

\begin{exbox}{Pre-Class Exercises (do BEFORE the lecture)}
\textbf{Exercise 3.1 --- Implement.} Complete the SLL methods (do not use a tail pointer):
\begin{lstlisting}
public void addFirst(E e);
public E    removeFirst();
public int  size();         // O(n) is fine
\end{lstlisting}

\textbf{Exercise 3.2 --- Why is removeLast slow on SLL?} In one sentence, explain why \texttt{removeLast} is $O(n)$ on a singly linked list \emph{even with a tail pointer}.

\textbf{Exercise 3.3 --- Reverse.} Write a method \verb|reverse()| that reverses an SLL in place in $O(n)$ time and $O(1)$ extra memory. Hint: walk the list with three pointers \texttt{prev}, \texttt{cur}, \texttt{next}.

\textbf{Exercise 3.4 --- DLL with sentinels.} Sketch a DLL with a header and trailer holding the elements [\,a, b, c\,]. Then show, on your sketch, the pointer changes for: (i) \texttt{addFirst(z)}, (ii) \texttt{remove(b)}.

\textbf{Exercise 3.5 --- Choose the right structure.}
\begin{enumerate}[label=(\alph*)]
  \item Browser back/forward navigation history.
  \item A list of pixels accessed by index for an image filter.
  \item A music playlist where users insert/remove songs anywhere.
  \item Storing the last 1000 log lines (newest pushes oldest out).
\end{enumerate}
For each, choose array, ArrayList, SLL, or DLL, and justify briefly.

\textbf{Exercise 3.6 --- Trace.} Starting from an empty DLL, draw the structure after:
\verb|addLast(1); addLast(2); addFirst(0); remove(1); addLast(3);|.
\end{exbox}

\begin{disbox}
\begin{itemize}[leftmargin=*]
  \item How does adding sentinel nodes eliminate edge cases? Show the bug if they are removed.
  \item When does the cache-friendliness of arrays beat the asymptotic advantage of linked lists?
  \item Discuss Exercise 3.3 --- in interviews this is the canonical linked-list question.
\end{itemize}
\end{disbox}

\end{document}
